Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-09-23 analysis/paper/paper.Rmd
Final edits before submission

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [@mapping2021; 30cm at the standard port Helgeroa in Larvik]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates. Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If this method is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact in this setting, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

# References

>>>>

The close association between Stone Age settlements in the northern parts of Scandinavia and shifting prehistoric shorelines was proposed at the end of the 19th century [@degeer1896], and was first applied as a dating method at the turn of the century [@brogger1905; @hollender1901]. Shoreline dating has been fundamental to Norwegian Stone Age archaeology ever since [e.g. @berg-hansen2009; @bjerck2008; @johansen1963; @bjerck1990; @breivik2014; @olsen1984; @nummedal1923; @mikkelsen1975; @mjærum2022; @shetelig1922; @solheim2018; @solheim2020b]. The method is used both independently, and to compliment other sources of temporal data such as typological indicators or radiometric dates. However, given the coarse and fuzzy resolution of established typological frameworks, the vast number of surveyed sites that only contain generic lithic debitage that could hail from any part of the period, and as the conditions for the preservation of organic material is typically poor in Norway, dating with reference to shoreline displacement is often the only and most precise method by which one can hope to date the sites. Shoreline dating is consequently fundamental to our understanding of the Norwegian Stone Age. This is both because it is central to the temporal framework on which our understanding of the period is based, but also because the method is only applicable so long as the societies in question have continuously settled on or close to the contemporary shoreline. Consequently, adherence or deviation from this pattern also has major implications for the socio-economic foundations of the societies in question.

@@

The shoreline displacement data used in this study are based on the so-called isolation basin method [e.g. @kjemperud1986; @romundset2011], which involves extracting cores from a series of basins situated on bedrock at different elevations beneath the marine limit, and dating the transition from marine to lacustrine sediments. Each basin thus represents a high precision sea-level index point (SLIP) which are combined in a continuous time series for RSL-change adjusted to a common shoreline isobase. The isobases are here contours indicating equal shoreline displacement over the same time span [@svendsen1987, 116]. To minimise the impact of variable uplift rates, the cored basins are located in as constrained of an area of the landscape as possible. Following from the morphology of the retreating ice sheet, the uplift is more stark towards the north-east, which needs to be adjusted for in the case that any basins are located any significant distance from the common isobase that runs perpendicular to this uplift gradient (Figure \@ref(fig:map-iso)). Furthermore, as the uplift has been greater immediately following the retreat of the ice, such adjustments, and thus potential uncertainty, will be more critical further back in time. The resulting SLIPs are most commonly interpreted as representing the isolation of the basins from the highest astronomical tide, which is adjusted to mean sea-level in the compilation of the displacement curves, based on the present-day tidal range. For simplicity, the tidal range is assumed to have been the same throughout the Holocene [@sørensen2014, 44]. The highest astronomical tide in the study area reaches around 30cm above mean sea-level [30cm at the standard port Helgeroa in Larvik, @mapping2021]. Furthermore, the confidence bands of the displacement curves and their trajectory are quite complex constructs as they are the integrated result of both expert knowledge and more objectively quantifiable parameters. The reason for this is that the curves do not only contain uncertainty as related to radiometric dates, which are well defined, but they are for example also dependent on the interpretation and analysis of sediment cores, the nature and condition of the basin outlets and the adjustment to a common isobase, as well as being based on expert knowledge of regional post-glacial geologic developments and local geomorphology, to name but a few factors [e.g. @romundset2011; @romundset2018; @svendsen1987; for an alternative approach see @creel2022]. For more details and evaluations done for the compilation of each curve, the reader is therefore referred to the individual publications.

@@

Although their accuracy is of course ultimately dependent on the veracity of the geological reconstruction, the high rate of RSL-change in this period does result in very precise dates. Above it was suggested that additional temporal data could be combined with the method to improve its accuracy and precision. Drawing on @jaksland2014, this example instead highlights the fact that the spatial nature of the method means that a consideration of the surrounding terrain and other sites can also help to increase the precision of the method if this can be used to exclude certain RSLs as unlikely for when a site was in use. One approach could also be to assess the spatial implication of a proposed shoreline date by simulating the adjusted sea-levels, as is done for Pauler 1 in Figure \@ref(fig:brunlanes)B, followed for example by a visual evaluation of the topography or by evaluating the distance and steepness of the slope to the shoreline. If this method is developed further, it could conceivably be possible to exclude certain elevations as unlikely for the position of the shoreline when the site was in use. Such approaches would make less of an impact for the Brunlanes sites, where the 95% HDR is already quite constrained, but could considerably improve the precision of the method in cases where RSL-change has been less severe (cf. Figure \@ref(fig:redate)).

@@

# References {-}

### 2022-09-23 analysis/supplementary-materials/supplementary.Rmd
Final edits before submission

The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the national heritage database Askeladden. The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29], but the extent of this is difficult to ascertain based on maps of the area. As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for these factors.

@@

The first archaeological finds from Rognlien were done in the 1880s and, following some smaller investigations, was excavated by @ingstad1970 in 1957. The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden. The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970, 26]. The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site. Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm. Steinar Solheim, also provided in @stokke2017, 24]. Given the 0m distance that the older dates give here, however, there does not appear to be grounds on which to reject the older dates based on sea-level change.

>>>>

The site limit for Auve was drawn manually based on the description in the publication of the site and the site geometry available in the national heritage database Askeladden [@directorate2018]. The site is located in and around an intersection of the road, parts of the area has been levelled and farmed, and an area to north-west of the site has reportedly been affected by land-slides [@ostmo2008, 16--29], but the extent of this is difficult to ascertain based on maps of the area. As the simulation results appear reasonable [see also map in @ostmo2008, fig.142], nothing was done to try to correct for these factors.

@@

The first archaeological finds from Rognlien were done in the 1880s and, following some smaller investigations, was excavated by @ingstad1970 in 1957. The site limit has been manually defined based on the description given in the publication and the spatial geometry of the site as defined in the database Askeladden [@directorate2018]. The manually defined limit could be slightly off, but should match the elevation of the site given in the publication [@ingstad1970, 26]. The ^14^C-dates from the original excavation was not seen as contemporaneous with the lithic material on the site. Burnt elk bone from the original excavation was therefore recently subjected to radiocarbon dating and is believed to match the occupation of the site [pers.comm. Steinar Solheim, also provided in @stokke2017, 24]. Given the 0m distance that the older dates give here, however, there does not appear to be grounds on which to reject the older dates based on sea-level change.
